Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-08-18 manuscript/revision2/revision_biorxiv.tex
revision2: journal neutral format for bioRxiv revision

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  9pt,
  twocolumn,
  twoside]{pnas-new}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% TODO: Add custom LaTeX header directives here
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Diverse Genotype-by-Weather Interactions in Switchgrass},
  pdfauthor={Alice H. MacQueen; Li Zhang; Samuel P. Smith; Jason E. Bonnette; Arvid R. Boe; Philip A. Fay; Felix B. Fritschi; David B. Lowry; Robert B. Mitchell; Francis M. Rouquette Jr; Yanqi Wu; Arbel Harpak; Thomas E. Juenger},
  pdfkeywords={allele-by-environment effect variation, rank-changing
genotype-by-environment interaction, photoperiod, cumulative
rainfall, genetic variation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\templatetype{pnasresearcharticle}

@@

\newcommand{\equalcont}{1}

\newcommand{\correspond}{2}

\author[a%
,\equalcont%
,\correspond%
]{Alice H. MacQueen}
\author[a%
,\equalcont%
%
]{Li Zhang}
\author[a%
,\equalcont%
%
]{Samuel P. Smith}
\author[a%
%
%
]{Jason E. Bonnette}
\author[b%
%
%
]{Arvid R. Boe}
\author[c%
%
%
]{Philip A. Fay}
\author[d%
%
%
]{Felix B. Fritschi}
\author[e%
%
%
]{David B. Lowry}
\author[f%
%
%
]{Robert B. Mitchell}
\author[g%
%
%
]{Francis M. Rouquette Jr}
\author[h%
%
%
]{Yanqi Wu}
\author[a%
%
%
]{Arbel Harpak}
\author[a%
%
,\correspond%
]{Thomas E. Juenger}

\affil[a]{University of Texas at Austin, Department of Integrative
Biology, Austin, 78712}
\affil[b]{South Dakota State University, Department of
Agronomy, Brookings, 57006}
\affil[c]{USDA-ARS, Grassland, Soil and Water Research
Laboratory, Temple, 76502}
\affil[d]{University of Missouri, Division of Plant
Sciences, Columbia, 65211}
\affil[e]{Michigan State University, Department of Plant Biology, East
Lansing, 48824}
\affil[f]{USDA-ARS, Wheat, Sorghum, and Forage Research
Unit, Lincoln, 68583}
\affil[g]{Texas A\&M University, Texas A\&M AgriLife Research and
Extension Center, Overton, 75684}
\affil[h]{Oklahoma State University, Department of Plant and Soil
Sciences, Stillwater, 74078}

\leadauthor{MacQueen}

\significancestatement{The timing of plant seasonal development
(phenology) has major impacts on fitness because of the steep fitness
cost of plant-environment mismatches. We infer the mixture of ways that
genetic effects on phenological traits depend on plant environment,
focusing on how effects covary with weather prior to the phenological
event (GxWeather). Most effects do not have GxWeather, but a minority
do. GxWeather is population-specific. The majority of effects on the
timing of vegetative growth in the Gulf population have GxWeather: these
effects differ in sign across environments and covary with a photoperiod
cue two weeks prior. A minority of effects on flowering date in the Gulf
and Midwest populations had GxWeather, covarying with a cumulative
rainfall and a photoperiod cue, respectively.}

\authorcontributions{T.E.J. designed research. D.B.L. contributed plant
material and resources. J.B., D.B.L., and T.E.J. designed and executed
field experiments. A.R.B., P.A.F., F.B.F., D.B.L., R.B.M., F.M.R., Y.W.,
and T.E.J. hosted field experiments. A.H.M., L.Z., and S.P.S. conducted
statistical and computational analyses with input from T.E.J. and A.H.
The manuscript was written by A.H.M. with contributions from all
authors.}

\authordeclaration{The authors declare no conflicts of interest.}

\equalauthors{\textsuperscript{\equalcont} A.H.M. contributed equally to
this work with L.Z. and S.P.S.}

\correspondingauthor{\textsuperscript{\correspond}To whom correspondence should be addressed. E-mail: alicem@uchicago.edu}
\correspondingauthor{\textsuperscript{\correspond}To whom correspondence should be addressed. E-mail: tjuenger@utexas.edu}

\keywords{allele-by-environment effect variation | rank-changing
genotype-by-environment interaction | photoperiod | cumulative
rainfall | genetic variation}

\begin{document}
\maketitle

\begin{abstract}
The timing of vegetative and reproductive growth in plants
(``phenological timings'') depend on genetic effects (G), environmental
(e.g., weather) cues, and their interaction. Here, we measure
phenological timings in two highly divergent switchgrass (\emph{Panicum
virgatum}) subpopulations using repeated plantings of cloned individuals
at eight sites across the central United States. The timing of
vegetative growth for the two subpopulations reversed between their two
natural ranges and had strong negative correlations between these
regions; in contrast, the timing of flowering was positively~correlated
between gardens. We expect that these phenotypic correlations consist of
polygenic effects on phenology which have distinct patterns of GxE
segregating at different mapped loci. Thus, we infer the mixture of ways
genetic effects impact phenological timings, such as across common
gardens (GxE) or with weather cues (GxWeather). We demonstrate that we
can identify genetic variation with GxWeather and assign genetic loci to
specific weather-based cues or other patterns. For example, in the Gulf
subpopulation, 65\% of genetic effects on the timing of vegetative
growth covary with daylength 14 days prior to green-up date, and 33\% of
genetic effects on the timing of flowering covary with cumulative
rainfall in the week prior to flowering. However, most variation in
genetic effects cannot be attributed to variation in weather variables.
Selective breeding for particular alleles at GxWeather loci could alter
flowering responsiveness in a photoperiod or rainfall-specific way. More
broadly, our approach refines the characterization of
genotype-by-environment interactions and can be implemented in any
species phenotyped in multiple environments.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
\firstpage[13]{2}

Plant phenological traits are important components of plant fitness that
are affected by multiple external environmental cues (e.g.~degree of
winter chilling, day length, temperature, soil fertility, and water
availability), which signal existing or upcoming growing conditions
(1--4). Genetic responses to environmental cues determine the speed,
timing, and energy apportioned to vegetative and reproductive growth and
shape plant physiological responses, lifespan, and lifetime production
of viable seed. Day length (or photoperiod) is one of the most
predictable environmental cues, and genetic sensitivity to photoperiod
protects plants from potentially fatal consequences of phenological
responses to temperature cues at the ``wrong'' time of year. However,
the utility of specific environmental cues depends on both features of
the environment, such as cue predictability and relevance, and the
species' adaptive strategies (5). Species with wide natural
distributions can have multiple distinct environmentally cued
phenological responses. For example, populations of sunflower
(\emph{Helianthus annuus}) exhibit day-neutral, facultative short day,
and facultative long-day flowering responses, which vary with their
environments (6, 7). These distinct genetic responses in different
environments are known as genotype-by-environment interactions, or GxE.

Flowering time, or the transition from vegetative to reproductive
growth, is a common subject of GxE research (6--13), a key output of
selection driving adaptation to local environments (2, 14--17), and a
selection target for crop improvement to adapt crops to local or future
environments (18). Changing flowering responsiveness to photoperiod cues
has allowed geographic range expansion and increased yields in several
cereal species (14, 19--23) and other crops (24, 25). Recent statistical
advances in studying phenological GxE have involved determining critical
environmental indices before the phenological event occurs, such as
photothermal time within a critical growth window (11). However, most
studies of flowering GxE focus on finding a single, best fitting form of
genotype-environment covariance, despite the key expectation that
different genetic subpopulations, and even different genomic regions,
have likely evolved distinct patterns of GxE (26). Additionally, despite
theoretical predictions that local adaptation should involve trade-offs
caused by antagonistic pleiotropy, or alleles with effects with opposing
fitness outcomes (27--30), previous empirical work has found limited
evidence of trade-offs caused by this form of GxE (15, 31, 32). However,
this work has been limited by a known statistical bias that reduced
detection of genetic effects that differ in sign (31, 33, 34). Thus,
despite substantial interest in the frequencies of various forms of GxE,
the frequency of sign-changing GxE relative to other forms of GxE
remains unknown.

Previous research suggests that switchgrass phenological timings should
have GxWeather and that these timings could differ by genetic
subpopulation. Switchgrass is considered a short-day plant with
reproductive development strongly linked to day of the year (35).
However, as part of its wide environmental adaptation across the eastern
half of North America, its photoperiodicity has been predicted to differ
by plant latitude of origin (36, 37). We previously found that divergent
Midwest and Gulf genetic subpopulations of switchgrass have distinct
sets of environmental adaptations associated with each of two fitness
proxies, biomass and overwinter survival (38). The Midwest genetic
subpopulation is primarily composed of individuals from the well-studied
upland switchgrass ecotype (39, 40), while the Gulf subpopulation has
individuals from the well-studied lowland ecotype and the phenotypically
intermediate coastal ecotype (38).

Here, we test for GxWeather for the timing of two phenological traits by
loading patterns of genetic effects on phenology at eight common gardens
onto many patterns of weather covariance at these gardens. To do this,
we phenotyped a diversity panel of hundreds of switchgrass genotypes
from the Midwest and Gulf subpopulations for the timing of vegetative
and reproductive development. We did this at eight common garden
locations spanning 17 degrees of latitude: these gardens covered the
majority of the latitudinal and climatic range of switchgrass and
captured the most comprehensive picture to date of the environmental
variation this species encounters. We defined multiple ways phenological
traits might covary with weather (Table 1) and additional ways
phenological traits might vary by garden (SI Appendix, Section S1), then
jointly re-estimated genetic effects on these timings at all eight
common gardens using the set of these covariance matrices that
significantly improved the modeled log-likelihood when included (SI
Appendix, Section S2-S5) (41). We used the Bayesian framework
\emph{mash} (multivariate adaptive shrinkage) developed by (41), to
refine effect size estimates from genome-wide association (GWAS).
\emph{Mash} allowed us to identify and specify multiple covariance
structures among genetic effect estimates across sites, including
structures that represent covariance in weather variables of interest.
Importantly, this method circumvented statistical biases in detecting
genetic effects with the same or opposite signs (42). To confirm our
genetic mapping of GxWeather, we compared genomic locations of the
significant posterior effect estimates from \emph{mash} to mapping
results from an outbred mapping population grown at the same sites. Our
analyses allowed us to describe the weather cues and types of GxE
affecting phenology in two divergent natural populations of switchgrass
across the species' latitudinal range.

@@

We expected patterns of GxE and GxWeather to vary at the locus,
subpopulation, and trait level. Thus, we were interested in modeling the
different ways polygenic effects might covary across our eight common
gardens(43). We jointly re-estimated the genetic effects of a subset of
SNPs across all eight common gardens using \emph{mash} models, which can
flexibly capture many modes of covariance across gardens (SI Appendix,
Sections S1-S5, Datasets 1-6). To obtain genetic effects, we used
subsets of effects from garden- and subpopulation GWAS (SI Appendix,
Section S2-S3; Table S1). To obtain covariance structures, we specified
both covariance structures introduced in the initial \emph{mash}
manuscript (e.g., garden-specific effects; equal effects at all gardens,
Figure~\ref{fig-covar} A), and GxWeather covariance matrices estimated
from the covariance of empirical weather patterns at each garden at
specific times before the phenological event (Table 1; SI Appendix,
Section S1). Then, we used a greedy algorithm to select covariance
matrices from each model's set that significantly improved the model
likelihood (SI Appendix, Section S4; Table S2). Finally, we fit
\emph{mash} models using the selected set of covariance matrices on a
subset of relatively unlinked (r\textsuperscript{2} \textless{} 0.2),
`strong' genetic effects with low \emph{p}-values (SI Appendix, Section
S5).

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
\documentclass[
  9pt,
  twocolumn,
  twoside]{simple-article}% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{flushend.sty}{\usepackage{flushend}}{}

\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{authblk}
\usepackage{orcidlink}
\definecolor{mypink}{RGB}{219, 48, 122}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{simple-article}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs

\hypersetup{
  pdftitle={Diverse Genotype-by-Weather Interactions in Switchgrass},
  pdfauthor={Alice H. MacQueen; Li Zhang; Samuel P. Smith; Jason Bonnette; Arvid R. Boe; Philip A. Fay; Felix B. Fritschi; David B. Lowry; Robert B. Mitchell; Francis M. Rouquette Jr; Yanqi Wu; Arbel Harpak; Thomas E. Juenger},
  colorlinks=true,
  linkcolor={Blue},
  filecolor={Blue},
  citecolor={Blue},
  urlcolor={Maroon},
  pdfcreator={LaTeX via pandoc}}

@@

\newif\ifnoAffil\noAffilfalse
\newif\ifoneAffil\oneAffilfalse

\newcommand{\setNumAffil}[1]{%
    \ifnum#1=0
        \noAffiltrue
    \else
        \ifnum#1=1
            \oneAffiltrue
        \fi
    \fi
}

\setNumAffil{8}

\ifnoAffil
%% No affiliation given....
\author{Alice H. MacQueen, Li Zhang, Samuel P. Smith, Jason
Bonnette, Arvid R. Boe, Philip A. Fay, Felix B. Fritschi, David B.
Lowry, Robert B. Mitchell, Francis M. Rouquette Jr, Yanqi Wu, Arbel
Harpak, Thomas E. Juenger}
\else
\ifoneAffil
%% Single affiliation given...
\author{Alice H. MacQueen}
\author{Li Zhang}
\author{Samuel P. Smith}
\author{Jason Bonnette}
\author{Arvid R. Boe}
\author{Philip A. Fay}
\author{Felix B. Fritschi}
\author{David B. Lowry}
\author{Robert B. Mitchell}
\author{Francis M. Rouquette Jr}
\author{Yanqi Wu}
\author{Arbel Harpak}
\author{Thomas E. Juenger}
\affil{University of Texas at Austin, Austin}
\else
%% Multiple affiliations given
\author[1]{Alice H. MacQueen}
\author[1]{Li Zhang}
\author[1]{Samuel P. Smith}
\author[1]{Jason Bonnette}
\author[2]{Arvid R. Boe}
\author[3]{Philip A. Fay}
\author[4]{Felix B. Fritschi}
\author[5]{David B. Lowry}
\author[6]{Robert B. Mitchell}
\author[7]{Francis M. Rouquette Jr}
\author[8]{Yanqi Wu}
\author[1]{Arbel Harpak}
\author[1]{Thomas E. Juenger}
\affil[1]{University of Texas at Austin, Austin}
\affil[2]{South Dakota State University, Brookings}
\affil[3]{USDA-ARS, Temple}
\affil[4]{University of Missouri, Columbia}
\affil[5]{Michigan State University, East Lansing}
\affil[6]{USDA-ARS, Lincoln}
\affil[7]{Texas A\&M University, Overton}
\affil[8]{Oklahoma State University, Stillwater}
\fi
\fi

\date{}

\fancyhead[C]{Alice H. MacQueen $\bullet$ Diverse Genotype-by-Weather
Interactions in Switchgrass}

\begin{document}

\renewcommand{\maketitlehookd}{%
\begin{abstract}
\noindent{}The timing of vegetative and reproductive growth in plants
(``phenological timings'') depend on genetic effects (G), environmental
(e.g., weather) cues, and their interaction. Here, we measure
phenological timings in two highly divergent switchgrass (\emph{Panicum
virgatum}) subpopulations using repeated plantings of cloned individuals
at eight sites across the central United States. The timing of
vegetative growth for the two subpopulations reversed between their two
natural ranges and had strong negative correlations between these
regions; in contrast, the timing of flowering was positively~correlated
between gardens. We expect that these phenotypic correlations consist of
polygenic effects on phenology which have distinct patterns of GxE
segregating at different mapped loci. Thus, we infer the mixture of ways
genetic effects impact phenological timings, such as across common
gardens (GxE) or with weather cues (GxWeather). We demonstrate that we
can identify genetic variation with GxWeather and assign genetic loci to
specific weather-based cues or other patterns. For example, in the Gulf
subpopulation, 65\% of genetic effects on the timing of vegetative
growth covary with daylength 14 days prior to green-up date, and 33\% of
genetic effects on the timing of flowering covary with cumulative
rainfall in the week prior to flowering. However, most variation in
genetic effects cannot be attributed to variation in weather variables.
Selective breeding for particular alleles at GxWeather loci could alter
flowering responsiveness in a photoperiod or rainfall-specific way. More
broadly, our approach refines the characterization of
genotype-by-environment interactions and can be implemented in any
species phenotyped in multiple environments.
\end{abstract}
}

\maketitle

\section{Significance Statement}\label{significance-statement}

The timing of plant seasonal development (phenology) has major impacts
on fitness because of the steep fitness cost of plant-environment
mismatches. We infer the mixture of ways that genetic effects on
phenological traits depend on plant environment, focusing on how effects
covary with weather prior to the phenological event (GxWeather). Most
effects do not have GxWeather, but a minority do. GxWeather is
population-specific. The majority of effects on the timing of vegetative
growth in the Gulf population have GxWeather: these effects differ in
sign across environments and covary with a photoperiod cue two weeks
prior. A minority of effects on flowering date in the Gulf and Midwest
populations had GxWeather, covarying with a cumulative rainfall and a
photoperiod cue, respectively.

\section{Introduction}\label{introduction}

Plant phenological traits are important components of plant fitness that
are affected by multiple external environmental cues (e.g.~degree of
winter chilling, day length, temperature, soil fertility, and water
availability), which signal existing or upcoming growing conditions
\citep{bauerle_photoperiodic_2012, andres2012genetic, korner2010phenology, maple2024}.
Genetic responses to environmental cues determine the speed, timing, and
energy apportioned to vegetative and reproductive growth and shape plant
physiological responses, lifespan, and lifetime production of viable
seed. Day length (or photoperiod) is one of the most predictable
environmental cues, and genetic sensitivity to photoperiod protects
plants from potentially fatal consequences of phenological responses to
temperature cues at the ``wrong'' time of year. However, the utility of
specific environmental cues depends on both features of the environment,
such as cue predictability and relevance, and the species' adaptive
strategies \citet{botero2015evolutionary}. Species with wide natural
distributions can have multiple distinct environmentally cued
phenological responses. For example, populations of sunflower
(\emph{Helianthus annuus}) exhibit day-neutral, facultative short day,
and facultative long-day flowering responses, which vary with their
environments \citep{blackman2013interacting, henry2014transitions}.
These distinct genetic responses in different environments are known as
genotype-by-environment interactions, or GxE.

Flowering time, or the transition from vegetative to reproductive
growth, is a common subject of GxE research
\citep{blackman2013interacting, henry2014transitions, aagren2017adaptive, brachi2010linkage, dittmar2014flowering, li2018genomic, romero2017study, bustos-korts2019},
a key output of selection driving adaptation to local environments
\citep{unterseer2016, andres2012genetic, wadgymar2017identifying, blumel2015flowering, fournier-level2016},
and a selection target for crop improvement to adapt crops to local or
future environments \citet{jung2009flowering}. Changing flowering
responsiveness to photoperiod cues has allowed geographic range
expansion and increased yields in several cereal species
\citep{unterseer2016, turner2005pseudo, faure2012mutation, hung2012zmcct, zakhrabekova2012induced, yang2013oself3}
and other crops \citep{pin2012multifaceted, weller2019parallel}. Recent
statistical advances in studying phenological GxE have involved
determining critical environmental indices before the phenological event
occurs, such as photothermal time within a critical growth window
\citet{li2018genomic}. However, most studies of flowering GxE focus on
finding a single, best fitting form of genotype-environment covariance,
despite the key expectation that different genetic subpopulations, and
even different genomic regions, have likely evolved distinct patterns of
GxE \citep{Weine2023.06.21.545998}. Additionally, despite theoretical
predictions that local adaptation should involve trade-offs caused by
antagonistic pleiotropy, or alleles with effects with opposing fitness
outcomes
\citep{levene1953genetic, felsenstein1976theoretical, kawecki2004conceptual, hedrick1986genetic},
previous empirical work has found limited evidence of trade-offs caused
by this form of GxE
\citep{des2013genotype, wadgymar2017identifying, millet2016}. However,
this work has been limited by a known statistical bias that reduced
detection of genetic effects that differ in sign
\citep{des2013genotype, anderson2013genetic, anderson2011evolutionary}.
Thus, despite substantial interest in the frequencies of various forms
of GxE, the frequency of sign-changing GxE relative to other forms of
GxE remains unknown.

Previous research suggests that switchgrass phenological timings should
have GxWeather and that these timings could differ by genetic
subpopulation. Switchgrass is considered a short-day plant with
reproductive development strongly linked to day of the year
\citet{mitchell1997predicting}. However, as part of its wide
environmental adaptation across the eastern half of North America, its
photoperiodicity has been predicted to differ by plant latitude of
origin \citep{parrish2005biology, casler2004latitudinal}. We previously
found that divergent Midwest and Gulf genetic subpopulations of
switchgrass have distinct sets of environmental adaptations associated
with each of two fitness proxies, biomass and overwinter survival
\citet{lovell2021}. The Midwest genetic subpopulation is primarily
composed of individuals from the well-studied upland switchgrass ecotype
\citep{porter1966analysis, milano2016genetic}, while the Gulf
subpopulation has individuals from the well-studied lowland ecotype and
the phenotypically intermediate coastal ecotype \citet{lovell2021}.

Here, we test for GxWeather for the timing of two phenological traits by
loading patterns of genetic effects on phenology at eight common gardens
onto many patterns of weather covariance at these gardens. To do this,
we phenotyped a diversity panel of hundreds of switchgrass genotypes
from the Midwest and Gulf subpopulations for the timing of vegetative
and reproductive development. We did this at eight common garden
locations spanning 17 degrees of latitude: these gardens covered the
majority of the latitudinal and climatic range of switchgrass and
captured the most comprehensive picture to date of the environmental
variation this species encounters. We defined multiple ways phenological
traits might covary with weather (Table 1) and additional ways
phenological traits might vary by garden (SI Appendix, Section S1), then
jointly re-estimated genetic effects on these timings at all eight
common gardens using the set of these covariance matrices that
significantly improved the modeled log-likelihood when included (SI
Appendix, Section S2-S5) \citet{urbut2019flexible}. We used the Bayesian
framework \emph{mash} (multivariate adaptive shrinkage) developed by
\citet{urbut2019flexible}, to refine effect size estimates from
genome-wide association (GWAS). \emph{Mash} allowed us to identify and
specify multiple covariance structures among genetic effect estimates
across sites, including structures that represent covariance in weather
variables of interest. Importantly, this method circumvented statistical
biases in detecting genetic effects with the same or opposite signs
\citet{10.1093/biostatistics/kxw041}. To confirm our genetic mapping of
GxWeather, we compared genomic locations of the significant posterior
effect estimates from \emph{mash} to mapping results from an outbred
mapping population grown at the same sites. Our analyses allowed us to
describe the weather cues and types of GxE affecting phenology in two
divergent natural populations of switchgrass across the species'
latitudinal range.

@@

We expected patterns of GxE and GxWeather to vary at the locus,
subpopulation, and trait level. Thus, we were interested in modeling the
different ways polygenic effects might covary across our eight common
gardens\citep{weine2024}. We jointly re-estimated the genetic effects of
a subset of SNPs across all eight common gardens using \emph{mash}
models, which can flexibly capture many modes of covariance across
gardens (SI Appendix, Sections S1-S5, Datasets 1-6). To obtain genetic
effects, we used subsets of effects from garden- and subpopulation GWAS
(SI Appendix, Section S2-S3; Table S1). To obtain covariance structures,
we specified both covariance structures introduced in the initial
\emph{mash} manuscript (e.g., garden-specific effects; equal effects at
all gardens, Figure~\ref{fig-covar} A), and GxWeather covariance
matrices estimated from the covariance of empirical weather patterns at
each garden at specific times before the phenological event (Table 1; SI
Appendix, Section S1). Then, we used a greedy algorithm to select
covariance matrices from each model's set that significantly improved
the model likelihood (SI Appendix, Section S4; Table S2). Finally, we
fit \emph{mash} models using the selected set of covariance matrices on
a subset of relatively unlinked (r\textsuperscript{2} \textless{} 0.2),
`strong' genetic effects with low \emph{p}-values (SI Appendix, Section
S5).

@@

\section{Author Contributions}\label{author-contributions}

T.E.J. designed research. D.B.L. contributed plant material and
resources. J.B., D.B.L., and T.E.J. designed and executed field
experiments. A.R.B., P.A.F., F.B.F., D.B.L., R.B.M., F.M.R., Y.W., and
T.E.J. hosted field experiments. A.H.M., L.Z., and S.P.S. conducted
statistical and computational analyses with input from T.E.J. and A.H.
The manuscript was written by A.H.M. with contributions from all
authors.

\section{Conflict of Interest}\label{conflict-of-interest}

The authors declare no conflicts of interest.

\section{Equal Contributor}\label{equal-contributor}

A.H.M. contributed equally to this work with L.Z. and S.P.S.

\section{Acknowledgements}\label{acknowledgements}
